\chapter{Introduction}

\section{Background \& Motivation}

% Background & motivation ~ why TAA and sorghum?
% \subsection{Sorghum and trans-aconitic acid}
\hspace{1em}Growing concerns over environmental sustainability and the finite nature of fossil fuel reserves have driven increasing interest in bio-based chemicals as alternatives to petroleum-derived compounds. The plastic manufacturing industry represents one sector particularly affected by this transition. Plasticizer is added to the polymer to enhance the flexibility of the plastic. The most commonly used plasticizer is made from synthetic compound such as di(2-ethylhexyl)phthalate (DEHP), a phthalic acid ester (PAE) that, despite its widespread industrial use, is recognised as chemically hazardous \cite{bruni_aconitic_2022}. trans-Aconitic acid (trans-aconitate; TAA) is a naturally occurring secondary metabolite accumulated in plant species including sugarcane (\textit{Saccharum officinarum}) and sorghum (\textit{Sorghum bicolor}), and has been proposed as a bio-based precursor for plasticizer synthesis, offering a sustainable alternative to synthetic PAEs \cite{bruni_aconitic_2022}.

\hspace{1em}Other than its uses as a plasticizer, TAA serves as a building block for bio-based polymer synthesis across a range of sectors \cite{bruni_aconitic_2022}, particularly in biomedical applications. It has been utilized as a scaffold constituent for bone tissue engineering, and has shown further promise in drug delivery systems, where its conjugation with chitosan microparticles modulates their biocompatibility and biodegradability to enhance therapeutic efficacy.

% \subsection{Natural sources of TAA}
\hspace{1em}TAA can be produced through microbial biosynthesis in metabolically engineered strains \cite{guo_efficient_2026}, or extracted directly from plant biomass, primarily from the stem juice of sugarcane and sweet sorghum \cite{bruni_aconitic_2022}. Among plant-based sources, sorghum is considered the more suitable feedstock on practical grounds. Sugarcane stem juice simultaneously serves as the principal substrate for refined sugar production and as an industrial fermentation feedstock, creating an inherent conflict between food and non-food uses. Sorghum presents no such constraint: its grain is harvested for food, while its stem, which carries no comparable nutritional value can be directed toward TAA extraction without competing with the food supply. This functional separation of plant parts positions sorghum as a practical and sustainable feedstock for TAA production.

% Role of TAA in plant physiology
% \subsection{Role of TAA in plant physiology}
\hspace{1em}Among the documented physiological roles of TAA in plants is its function as a stable secondary carbon pool derived from the TCA cycle. Under active photosynthesis, the TCA cycle operates as an open, non-cyclic structure rather than a closed loop, leading to the accumulation of organic acid intermediates, including TAA, that would otherwise be rapidly consumed \cite{igamberdiev_organic_2016, igamberdiev_citrate_2020, igamberdiev_mitochondria_2023}. TAA itself arises from cis-aconitate, a TCA cycle intermediate, via isomerization to its chemically stable trans-form \cite{igamberdiev_organic_2016}. Its stability allows it to accumulate to significant concentrations in plant tissues, and its biosynthesis has been shown to be coupled to light availability, suggesting a functional link between TAA metabolism and high photosynthetic activity \cite{igamberdiev_organic_2016}.

\hspace{1em}Additionally, TAA also functions as a chemical defense compound in plants. In wheat, TAA serves as a precursor to methylated trans-aconitate derivatives that accumulate at sites of fungal penetration by \textit{Blumeria graminis} and exhibit potent antifungal activity, with their production further amplified by silicon treatment, consistent with a phytoalexin-like role. The conversion of TAA to its antifungal methylated forms is proposed to be catalyzed by a trans-aconitate \textit{O}-methyltransferase \cite{remus-borel_aconitate_2009}; however, no such enzyme has been characterized in plants to date. The silicon-amplified accumulation of methyl-TAA further suggests that this methylation step is subject to stress-inducible regulation, adding a layer of enzymatic control to the defense response \cite{remus-borel_aconitate_2009}. Against insect herbivores, TAA stored in the non-phloem tissues of barnyard grass (\textit{Echinochloa crus-galli}) deters feeding by the brown planthopper \textit{Nilaparvata lugens}: upon probing through parenchyma cells, the insect encounters TAA and withdraws its stylets before reaching the phloem, effectively preventing nutrient ingestion \cite{hattori_probing_2001}. TAA also exerts allelopathic effects in the soil environment. When released by TAA-accumulating grasses into rotation systems, it is absorbed by neighboring plant roots, where it induces reactive oxygen species overproduction, increases membrane permeability, impairs water and nutrient uptake, and ultimately suppresses growth \cite{bortolo_trans-aconitic_2018}.

\section{Research context}
\hspace{1em}Despite the established roles of TAA in plant metabolism and defense, the genes encoding the enzymes responsible for its biosynthesis remain unidentified in any plant species. The primary biosynthetic enzyme, aconitate isomerase (EC 5.3.3.7), has been detected and partially purified in maize. Primers designed from conserved regions of known bacterial and fungal aconitate isomerase sequences were used to detect the enzyme in the cytosol of the leaves \cite{eprintsev_aconitate_2021}. Expression was found to be light-dependent and responsive to red light, suggesting regulation by the phytochrome system. This is consistent with the broader link between TAA accumulation and photosynthetic activity \cite{eprintsev_aconitate_2021, igamberdiev_organic_2016}. Despite this progress, the responsible gene was not concretely reported in the paper.

\hspace{1em}Two additional gaps remain in connecting TAA to its defense functions. First, since TAA is produced as a photosynthetic overflow metabolite, leaves are the likely primary site of biosynthesis; yet no transporter has been described that moves TAA to storage sites such as pith parenchyma vacuoles, nor has the accumulation of TAA specifically in non-phloem tissues of barnyard grass been attributed to any known carrier. Second, \cite{remus-borel_aconitate_2009} showed that TAA must be methylated to function as an antifungal phytoalexin in wheat \cite{remus-borel_aconitate_2009}, but the plant enzyme responsible for this conversion has not been identified and its expression pattern under pathogen stress is unknown.

% transitioning to research objective
\section{Research Objective}
\hspace{1em}Closing these gaps requires a strategy that does not rely on sequence similarity to the characterized bacterial and fungal orthologs \cite{eprintsev_aconitate_2021}. Forward genomics offers such a route: by exploiting the natural variation in TAA accumulation and secretion across \textit{Sorghum bicolor} accessions, candidate genes can be prioritized directly from genotype--phenotype associations rather than from homology. To this end, this thesis constructs a multi-omics platform that integrates genomic, epigenomic, and transcriptomic evidence into a ranked list of candidate genes underlying metabolic variation in sorghum. TAA biosynthesis and secretion serve as the primary application of the platform. In parallel, the D-gene, a well-characterized transcription factor controlling stem juiciness \cite{fujimoto_transcriptional_2018}, serves as a positive control: recovering this known gene validates that the platform behaves as intended before it is trusted on the unresolved TAA pathway.

\section{Methods Overview}
\hspace{1em}The platform integrates three omics layers, all anchored on four \textit{Sorghum bicolor} accessions of contrasting phenotypes sequenced with Oxford Nanopore (ONT) long-read sequencing. At the genomic layer, variants are called for each accession against the reference genome, yielding accession-specific variant sites. At the epigenomic layer, DNA methylation is profiled and used to identify differentially methylated regions (DMRs) between each pair of accessions, producing six pairwise comparisons. At the transcriptomic layer, a gene co-expression network is constructed following the ATTED-II framework from publicly available expression data. % Evidence from these three layers is then combined to rank candidate genes, with the goal of identifying the gene encoding aconitate isomerase --- the first committed step of TAA biosynthesis --- in \textit{Sorghum bicolor}.

