\chapter{Methods}

% about the platform (elaborating the three -omics layer and its example usage)
% the omics layers
\section{Construction of the Platform}
\hspace{1em}The data for the platform is generated through variant analysis, DMR analysis, and gene coexpression analysis, each represents the genomics, epigenomics, and transcriptomics layer, respectively. The genomics and epigenomics layers are constructed from four sorghum strains exhibiting different phenotypes, collectively referred to as "TAA-research strains". For genomics, we performed variant analysis to obtain variant sites, assigned these variants to genes, then predicted whether the variant could potentially alter the phenotype associated. For epigenomics, we conducted differentially methylated region (DMR) analysis to obtain DMRs, associate these DMRs with genes' promoters, and determine which genes are transcriptionally repressed due to the methylated promoters. Each from these two approaches, a set of prioritized candidate genes associated with certain phenotypes are obtained. 

% gene co-ex layer
\hspace{1em}The gene coexpression analysis data is generated from a different set of data. Gene coexpression relationships were calculated following the framework of the ATTED-II database \cite{obayashiATTEDIIV11Plant2022}, using publicly available \textit{Sorghum bicolor} gene expression (RNA-seq) data. The analyses that are based on the gene coexpression data mainly resolve on the elucidation of the existing knowledge base of a certain gene network cluster related to the phenotype of interest.

% demonstration or somewhat a validation
\section{Platform Demonstration and Application}
\subsection{Stem juiciness}
\hspace{1em}For demonstration, the two phenotypes presented in this study are stem juiciness and TAA concentration in stem juice. Stem juiciness is chosen as the positive control since the metabolic pathway and enzymatic machinery that controls this phenotype is already characterized and identified, although partially. Stem juiciness is occurred when the pith parenchyma tissues of the stem remains alive and not undergoes a developmental programmed cell death (dPCD). Consequently, any disruption in the expression pathway of genes in the dPCD module would cause stem juiciness trait to arise. The constituent of the dPCD module, consisting of transcription factor genes (NAC family) and PCD executor enzymes, i.e., nucleases, peptidases, etc., was reported, although not in sorghum, but model organism \textit{Arabidopsis thaliana} (Appendix~\ref{tab:dpcd-broad}) \cite{olvera-carrilloConservedCorePCD2015}. In addition, stem juiciness is also an important agricultural trait in sorghum, notably for the TAA extraction process from its stem juice.

% demonstration but this one is more like a further application
\subsection{TAA metabolic pathway}
\hspace{1em}At the current state, there is still no record for the identification of genes working in the TAA metabolism pathway, let alone its biosynthesis. However, one prominent example of TAA-related enzyme, aconitate isomerase, has been studied and found in maize \cite{eprintsevAconitateIsomeraseMaize2021}. Aconitate isomerase catalyzes the conversion of \textit{cis}- and \textit{trans}- isomers of aconitic acid in a reversible reaction. The expression of this enzyme is induced by white and by red light, indicating that phytochrome is involved in its regulation expression. On the other hand, a set of respiratory enzymes whose expression is also induced by light and upregulated by phytochrome is already documented in \textit{A. thaliana} (Appendix~\ref{tab:phy-module}) \cite{igamberdievMitochondriaPhotosyntheticCells2023} \cite{teppermanExpressionProfilingPhyB2004} \cite{escobarLightRegulationArabidopsis2004}. Analyzing coexpressed genes in this module through gene coexpression analysis might aid the identification attempt of aconitate isomerase candidates.

\section{Plant Material and Sampling}
\hspace{1em}Four \textit{Sorghum bicolor} accessions with contrasting phenotypes were analysed in this study: SBC4, SBC10, SBC11, and SBC23, collectively referred to as "TAA-research strain". The accessions differ in their TAA concentration in stem juice and juice production capacity (Table~\ref{tab:samples}). The genomic DNA of the four sample was extracted from leaf tissue of each accession. The BTx623 inbred line (\texttt{GCF\_000003195.3} assembly \texttt{Sorghum\_bicolor\_NCBIv3}) served as the reference genome for all downstream analyses. The reference sample is known to possess a juicy phenotype \cite{zhangSweetSorghumOriginated2018} with assumed average TAA concentration in the juice.

\begin{table*}[h]
    \centering
    \caption{The four \textit{Sorghum bicolor} accessions analysed in this study and their contrasting phenotypes.}
    \label{tab:samples}
    \begin{tabularx}{\textwidth}{c >{\centering\arraybackslash}X >{\centering\arraybackslash}X >{\centering\arraybackslash}X}
        \toprule
        Sample & Cultivar & TAA conc. in juice & Juice production \\
        \midrule
        Reference & BTx623 & + & + \\
        SBC4 & IS32787 & ++ & ++ \\
        SBC10 & IS20956 & -- & +++ \\
        SBC11 & S. vulgare 72-726-7 & ++ & -- \\
        SBC23 & 240 Wad Umm Benein & ++ & ++ \\
        \bottomrule
    \end{tabularx}
\end{table*}

\section{Sequencing}
\hspace{1em}Genomic DNA from each accession was sequenced on the Oxford Nanopore Technologies (ONT) R10.4.1 platform in duplex mode at 400~bps. Raw signal data were basecalled with Dorado v1.3.0 using the super-accurate (SUP) duplex model, with modified-base calling enabled (\texttt{--modified-bases}) to retain 5-methylcytosine (5mC) information in the \texttt{MM}/\texttt{ML} BAM tags required for downstream methylation analysis.

\hspace{1em}Basecalled reads were aligned to the BTx623 reference genome with minimap2 v2.30 \cite{liMinimap2PairwiseAlignment2018}, and the resulting alignments were coordinate-sorted and indexed with samtools v1.21 \cite{liSequenceAlignmentMap2009}. SBC4, SBC10, and SBC23 were each derived from a single sequencing library, whereas SBC11 was sequenced across three libraries that were merged with \texttt{samtools merge} and re-sorted prior to analysis. Per-position read depth was computed with \texttt{samtools depth} for each accession (Appendix \ref{fig:depth-sbc4} - \ref{fig:depth-sbc23}).

\section{Variant Calling and Annotation}
\hspace{1em}Small variants (SNVs and indels) were called for each accession against the reference genome using Clair3 \cite{zhengSymphonizingPileupFullalignment2022} with the ONT R10.4.1 super-accurate (SUP) model, yielding accession-specific variant call sets. Raw calls were filtered with bcftools v1.21 to retain high-confidence variants (\texttt{PASS} filter only; QUAL $\geq$ 20, read depth between 10 and 100) located on the ten nuclear chromosomes and the mitochondrial and chloroplast contigs, and the filtered variants were phased into haplotypes with WhatsHap v2.8. The four accessions' phased VCFs were then merged into a single multi-sample VCF in canonical sample order with bcftools, splitting multiallelic sites and left-aligning indels so that each alternate allele forms its own record; this merged VCF is the pipeline's single source of truth, with any accession-sharing decomposition treated as a downstream analysis step rather than part of the annotation pipeline itself.

\hspace{1em}Structural variants (SVs), such as large deletions, insertions, duplications, inversions, and breakends not captured by Clair3, were called for each accession from the same aligned BAMs used for small-variant calling, using Sniffles2 \cite{smolkaDetectionMosaicPopulationlevel2024}. Each accession's SV calls were produced alongside a \texttt{.snf} snapshot file, which Sniffles2 uses for population-scale merging; the four accessions' \texttt{.snf} files were then force-genotyped jointly into a single multi-sample SV VCF. This population-merge mechanism was used in place of a positional VCF merge, whose exact POS/REF/ALT matching fragments SV breakpoints across samples that Sniffles2's own merge mode resolves correctly. Structural variants of $\geq$100~kbp in absolute length, predominantly whole-chromosome-arm duplications and inversions arising from long-read mapping noise across repetitive regions, were removed from the combined SV VCF prior to annotation.

\hspace{1em}The merged SNV/indel VCF was annotated once for functional consequence with SnpEff \cite{cingolaniProgramAnnotatingPredicting2012b}, using a custom \textit{Sorghum bicolor} (NCBIv3) database built from the NCBI gene annotation (\texttt{GCF\_000003195.3} GFF3 file); because SnpEff's predictions are per-allele rather than per-sample, a single run assigns each variant a predicted effect (e.g.\ missense, synonymous, or stop-gain) and impact category while retaining all four accessions' genotype columns. Because the VCF contig identifiers (e.g.\ \texttt{NC\_012870.2}) differ from the numeric chromosome names used by SnpEff, a chromosome synonym mapping was applied to reconcile the two naming schemes. The SnpEff-annotated multi-sample VCF was then scored once with SIFT4G \cite{vaserSIFTMissensePredictions2016a} against a \textit{Sorghum bicolor} (NCBIv3) prediction database built from the reference genome and gene models, using UniRef90 as the homology protein database, classifying each coding substitution as tolerated or deleterious. In particular SIFT4G annotates the impact of missense variant, with regard to the conserved domain information from the said protein database. The filtered, combined SV VCF was likewise annotated once for functional consequence with the same SnpEff \textit{Sorghum bicolor} database and chromosome synonym mapping used for the SNV/indel VCF, so that gene identifiers are directly comparable between the two variant classes; SIFT4G was not applied to SVs, as its coding-substitution model does not extend to large structural changes.

\hspace{1em}Finally, the SIFT4G-annotated SNV/indel VCF and the SnpEff-annotated SV VCF, both retaining the same four accessions in canonical order, were concatenated with bcftools into a single multi-sample VCF carrying both variant classes, which serves as the common input for all downstream variant-based analyses.

\section{DMR Calling and Annotation}
\hspace{1em}Per-accession 5mC methylation was quantified from the \texttt{MM}/\texttt{ML}-tagged alignments using modkit v0.2.6, with the presence of modified-base tags verified beforehand using \texttt{modkit summary}. For the genome-wide methylation landscape (Figure~\ref{fig:context-methylation-level}), bedMethyl pileups were generated separately for the three plant sequence contexts (CpG, CHG, and CHH) and summarised as mean methylation per context. The differentially methylated region (DMR) analysis used the CpG context only: CpG bedMethyl pileups with strands combined were produced, and positions covered by at least ten valid reads were retained for differential analysis.

\hspace{1em}Differentially methylated regions (DMRs) were identified with the DSS R package \cite{fengDifferentialMethylationAnalysis2019} from the filtered 5mC profiles, restricted to the ten main sorghum chromosomes. DMRs were called for each pairwise comparison of accessions, yielding six comparisons in total, which were grouped according to the TAA phenotype contrast. For each comparison, differential methylation was tested at single-CpG resolution with \texttt{DMLtest} using smoothing (\texttt{smoothing.span}~$=500$), and DMRs were subsequently called with \texttt{callDMR} requiring a minimum methylation difference of 0.1, a p-value threshold of 0.2, a minimum region length of 50~bp, and at least 20 CpG sites per region; neighbouring regions separated by less than 300~bp were merged. The relaxed p-value threshold reflects the exploratory, single-replicate design ($n=1$ per accession), and the resulting DMRs were treated as exploratory rather than statistically definitive. The per-comparison DMR tables were combined and annotated against the BTx623 gene models (NCBIv3 GFF) by overlap with four genomic feature types: promoters, exons, CDS, and introns. Promoters were defined in a strand-aware manner as the 2~kbp region immediately upstream of each gene's transcription start site. Each DMR was assigned to every feature it overlapped, with overlaps resolved against all annotated genes; DMRs overlapping no feature were labelled intergenic.

\section{Coexpression data}
\hspace{1em} Gene coexpression relationships were obtained following the framework of the ATTED-II database \cite{obayashiATTEDIIV11Plant2022}, using publicly available \textit{Sorghum bicolor} gene expression data. The sorghum coexpression network was constructed from the expression patterns of 21,627 genes across 807 publicly available RNA sequencing samples, calculated using the same method applied across the ATTED-II framework, and was published online as part of the ATTED-II v13.0 release (\url{https://www.atted.jp}).

\hspace{1em}The resulting network was evaluated against the KEGG pathway database (data downloaded 15 December 2025), restricted to the 1{,}337 genes present in both the KEGG annotation and the sorghum coexpression data. This evaluation yielded a score of 3.927, comparable to other species hosted on ATTED-II (\textit{Arabidopsis thaliana}: 7.093; \textit{Zea mays}: 5.206; \textit{Nicotiana tabacum}: 3.569; random: 0.500)


