\chapter{Discussion}

% recap
\hspace{1em}In this study, we constructed a multi-omics platform to help identify the genetic basis of metabolic variation in \textit{Sorghum bicolor}, combining variant analysis, promoter methylation, and gene coexpression analysis, and asked whether it could recover known biology before applying it to an uncharacterised trait. Stem juiciness served as the positive control, testing whether the platform re-identifies the dPCD gene module that dictates pith parenchyma cell death; the TAA case then tested its utility where no ground truth exists. Together, the two cases show what the platform achieves and how its results should be read.

\section{Construction of the platform}
% demonstration on dPCD gene module discussion
\hspace{1em}The demonstration shows that the platform produces only a weak, nominal enrichment of the dPCD module, and only in one of the four enrichment tests. The variant layer for the stem extra juiciness phenotype was the only test with a nominal enrichment (1.27-fold, uncorrected $p = 0.041$); the corresponding dryness contrast was enriched in the same direction but more weakly, and no methylation-derived list showed appreciable over-representation. Because the four tests were run without multiple-testing correction, even this nominal enrichment should be read cautiously. The scale of the candidate lists explains why the enrichment is so faint: the 13,718 variant candidates for extra juiciness already represent roughly 42\% of the 32,945-gene background, so recovering a comparable fraction of any gene set is the null expectation.

\hspace{1em}Two factors likely contribute to this weak enrichment. The first is the experimental design and the reference module it is measured against, both discussed below. The second is biological: stem juiciness in sorghum may be governed in part by genes lying outside the dPCD module, most notably the \textit{Dry} NAC transcription factor gene (EGI: 8084661) \cite{fujimotoTranscriptionalSwitchProgrammed2018}, so a module defined purely from conserved PCD markers is not expected to capture the full genetic basis of the trait.

% application on TAA metabolism pathway
\hspace{1em}Applied to the TAA pathway, the platform produces hypotheses rather than confirmed candidates. Because no gene of the TAA pathway has been identified in any plant species, let alone \textit{Sorghum bicolor}, the ORA framework used for the demonstration could not be applied here. The variant and methylation layers still narrowed the genome to 3,032 and 2,617 candidate genes respectively, and the search for aconitate isomerase, the key enzyme of the pathway, fell to the coexpression analysis. Its candidates are drawn from a broad Gene Ontology "isomerase" annotation spanning 461 genes, and the strongest relationship it finds ( between heat shock protein 82 and a 70~kDa peptidyl-prolyl isomerase ($z = 8.88$)) links a light-responsive chaperone to a protein-folding enzyme with no known role in organic-acid interconversion.

% about integration in this platform
\hspace{1em}The integration here is essentially additive: each layer contributes a separate line of evidence pointing to a different possible cause. The genomics layer marks genes altered by high- or moderate-impact variants, the epigenomics layer marks genes with differentially methylated promoters, and the transcriptomics layer places these candidates within the coexpression structure of the genome. Because each captures a different biological mechanism, adding a layer does not refine a single estimate but introduces a further, complementary hypothesis, so a gene supported across several layers is a stronger candidate than one flagged by only one.

% limitation
\section{Limitations}
\hspace{1em}Several limitations should be kept in mind when reading these results. The first is the size of the experiment: each accession is represented by a single plant ($n = 1$), and both phenotypes are recorded as broad, ordinal categories (juicy or dry, high or low TAA) rather than measured on a continuous scale. With so few data points, a real causal difference cannot be separated confidently from the ordinary variation between individual plants. The second is the length of the candidate lists. The variant layer nominates 13,718 genes for stem extra juiciness, roughly 42\% of the 32,945-gene background, so a list this long will recover a similar share of any gene set by chance alone. The nominal enrichment sits on top of a very permissive filter.

\hspace{1em}A third limitation is the reference set used to validate the platform. The dPCD module was curated in \textit{Arabidopsis thaliana} and carried into sorghum by orthology, and the coexpression evidence comes from public ATTED-II data rather than from the four accessions studied here, so it describes how genes behave in plants generally, not in these particular plants. For TAA there is no reference set at all, so its candidate lists cannot be checked in the same way.

% future direction
\section{Future direction}
\hspace{1em}The most direct improvement is a larger and better-measured panel. Both case studies compare one accession against a small group of others using a binary phenotype call, so each candidate list rests on very few data points. More accessions per phenotype class, with the phenotype measured on a continuous scale, would allow the trait to be mapped statistically through genome-wide association studies (GWAS), which draw their power from many diverse accessions rather than from a handful of contrasts; a quantitative trait locus (QTL) mapping approach would instead require a purpose-built mapping population, such as a biparental cross. In practice this means recording TAA as an actual concentration across many accessions instead of sorting a few into high and low bins, and scoring stem juiciness by the degree of aerenchyma formation and pith parenchyma cell death, since that is the developmental process the D-gene actually controls. Both changes would replace the coarse comparisons used here with a properly powered mapping study, giving every layer a stronger and more precise signal.

\hspace{1em}The platform's reference data could be strengthened in parallel. A dedicated study of stem juiciness in sorghum would allow a sorghum-specific dPCD module to be assembled, giving the enrichment test a reference set defined in the right species and the coexpression analysis query genes that actually act in sorghum. Measuring expression directly in these four accessions, rather than relying on ATTED-II, would tie the coexpression layer to the same plants as the other two, and protein--protein interaction data would help separate genes that are merely coexpressed from those likely to interact. For TAA, the candidate lists are best treated as a starting point for laboratory work: the highest-ranked isomerase genes can be tested through heterologous expression or knockout, the only way to convert a prioritised hypothesis into an identification.

% conclusion
\section{Conclusion}
\hspace{1em}This study set out to build a multi-omics platform for sorghum and to ask whether it could recover known biology before being applied to an unknown trait. The answer from the positive control is a qualified yes: the platform produced only a weak, nominal enrichment of the dPCD module, in one of four enrichment tests (1.27-fold, uncorrected $p = 0.041$), and it did not recover the D-gene itself. Tracing why is what makes the control useful. The misses are consistent with the design and the reference sets --- a single plant per accession, a coarse phenotype, very long candidate lists, and a module curated in another species that excluded the D-gene by construction.

\hspace{1em}Applied to TAA, where no ground truth exists, the platform narrowed the genome to 3,032 variant-derived and 2,617 methylation-derived candidate genes, and ranked 461 annotated isomerase genes by their coexpression with a phytochrome-upregulated proxy module. None of the top-ranked candidates carries an annotation linked to organic-acid metabolism, so aconitate isomerase is not identified here. A systematic cross-layer comparison --- testing whether the variant-, methylation-, and coexpression-derived candidate lists converge on the same genes --- was beyond the scope of this study and is left as an important direction for future work. As they stand, these lists are what the platform is designed to produce: prioritised, testable hypotheses that narrow the search space for targeted follow-up. Taken together, the two case studies show that multi-omics evidence can be connected to candidate genes in sorghum, and, just as importantly, they show where that evidence is currently thin and what a better-designed study would need to fix.

% \section{Obtaining D-gene using the platform}
% \subsection{Causative gene candidates list from variant analysis}
% \hspace{1em}The variant layer showed a modest but statistically significant rank enrichment of the dPCD module (Mann--Whitney $p = 0.015$), with two cutoff-based over-representation checks pointing in the same direction without reaching significance on their own. This result, however, describes only an aggregate, module-wide trend: the 65 testable dPCD executor orthologs collectively skew toward higher damage scores than the rest of the scored gene set, which is a weaker and different claim than saying that any specific module gene -- let alone the true causative D-gene itself -- was individually prioritised.

% \hspace{1em}Checking the true D-gene directly makes this distinction concrete. Fujimoto et al.\ \cite{fujimoto_transcriptional_2018} identify the sorghum D-gene as Sobic.006G147400 (LOC8084661 in the NCBI annotation used throughout this thesis), the ortholog of \textit{Arabidopsis} \textit{ANAC074}. This gene is nominally present in the candidate variant pool -- 575 variant--gene rows survive the SBC11-exclusion filter for it -- but its single most severe annotated consequence across the four accessions is a MODIFIER-impact \texttt{upstream\_gene\_variant}, the lowest severity band in the scoring scheme. Because gene-level scoring takes the maximum (worst) score across a gene's variants, LOC8084661 is pinned near the bottom of the ranked gene list despite technically clearing the initial site-selection filter. A structural-variant check at the same locus found no clean juicy-private candidate either: SVs overlapping LOC8084661 appear in several accession-sharing groups, including all four accessions together, but not in the SBC4\_SBC10\_SBC23 (juicy-only) group that would mirror the SNP/indel selection logic. Across every genomic track available in this four-accession panel, no variant offers a clean, phenotype-discriminating signal at the true D-gene locus.

% \hspace{1em}This does not contradict the module-level enrichment result -- it illustrates that enrichment and single-gene recovery are separate claims that this analysis can satisfy independently of one another. A damaging-variant-based search will only surface a gene if its own coding sequence carries a qualifying lesion in this specific panel; if the causal difference instead lies in a regulatory element, a structural rearrangement not well captured by SNV/indel calling, or simply an allele not present among these four accessions (Fujimoto et al.\ report six independent naturally occurring nonfunctional D alleles across a broader sorghum diversity panel), a coding-impact score will not detect it.

% \subsection{Causative gene candidates list from DMR analysis}
% \hspace{1em}The DMR layer showed no enrichment of the dPCD module (hypergeometric ORA, fold $= 0.92$, $p = 0.836$), and the true D-gene offers a clean explanation for this null result rather than an unexplained failure. \texttt{LOC8084661} has zero DMR-feature rows anywhere in the annotated DMR table: DSS never called a differentially methylated region overlapping this locus in any of the six pairwise comparisons, so the gene was never even eligible to enter the DMR candidate pool, independent of the hypomethylation-in-SBC11 selection criterion applied afterwards.

% \hspace{1em}This is consistent with the D-gene's known molecular basis. Fujimoto et al.\ report that the juicy phenotype in natural sorghum populations arises from independent nonfunctional \textit{D} alleles -- a genetic loss-of-function mechanism, not the epigenetic silencing of an otherwise intact wild-type allele. Promoter-methylation-based candidate selection is designed to detect the latter: a gene whose regulatory region has been differentially silenced between accessions that otherwise retain a functional coding sequence. Since the D locus's causal variation is allelic rather than epigenetic, a DMR-based search was targeting a molecular mechanism that does not apply at this specific locus, and the resulting null result is better read as a mismatch between the platform's epigenetic-silencing hypothesis and the D-gene's actual biology than as a failure of the DMR pipeline itself.

% \subsection{D-gene transcription factor from the coexpression analysis}
% \hspace{1em}The coexpression layer ranked the five NAC-family transcription factors present in the 66-gene dPCD module by how consistently each was coexpressed with the module's other 61 genes, nominating LOC8080837 (\textit{protein NTM1-like 9}, chromosome 7) as the most probable D-gene candidate. This is not the true D-gene: LOC8084661 sits on chromosome 6, matching Fujimoto et al.'s fine-mapping interval, and -- more fundamentally -- it is not a member of the 66-gene dPCD module at all. That module was built from orthologs of the conserved dPCD \textit{executor} gene families characterised by Olvera-Carrillo et al.\ \cite{olvera-carrillo_conserved_2015} (ribonucleases, serine carboxypeptidases, aspartic proteinases), whereas D acts upstream of these executors as their transcriptional switch, placing it outside the executor orthology group the module was curated from.

% \hspace{1em}The coexpression search was therefore structurally unable to nominate the true D-gene, regardless of how well the ranking method performed among the five candidates it did consider: a guilt-by-association search that ranks a module's own members can only succeed if the true regulator is already a latent member of that module, and here it was excluded by the module's construction before any coexpression computation took place. This traces the platform's stated limitation for this case study -- failing to recover the D-gene via coexpression -- to a specific, identifiable cause: the module curation step, not the coexpression data or the ranking methodology, was the bottleneck.

% \section{Obtaining aconitate isomerase using the platform}
% \subsection{Causative gene candidates list}
% \hspace{1em}Unlike the D-gene case study, aconitate isomerase has no independent ground truth to check candidate recovery against: the enzyme (EC 5.3.3.7) remains molecularly unidentified in any plant species, so there is no known gene analogous to LOC8084661 to search for among the variant-derived (21,928 scored genes) or DMR-derived (2,617 candidate genes) lists. In this setting the platform's output is necessarily an unvalidated hypothesis-generation exercise rather than a check against a known answer, and its credibility depends on internal consistency -- for instance, whether independent layers converge on the same genes -- rather than on recovering an established target.

% \hspace{1em}Checking for such convergence, none of the coexpression layer's top five candidates (Table~\ref{tab:taa-coexpression-ranking}) appear among the DMR layer's 2{,}617 promoter-hypomethylated candidates. The absence of overlap is not, on its own, strong evidence against any individual layer: with candidate lists this large relative to the roughly 35,000-gene sorghum genome, substantial overlap would require real biological convergence to be likely by chance, and its absence here only weakens the additional claim that the layers agree with one another, not the validity of each layer in isolation.

% \subsection{Identification of aconitate isomerase from the coexpression analysis}
% \hspace{1em}The three-gene phytochrome-upregulated proxy module used to rank isomerase-family candidates was itself flagged, at the point it was constructed, as a lower-confidence substitute for a true literature-curated module: it reflects convergence on a qualitatively similar light-induction pattern across heterologous \textit{Arabidopsis} experiments rather than coexpression measured within a single shared dataset, unlike the dPCD module used for the D-gene case study. Any ranking built on it inherits that same uncertainty.

% \hspace{1em}Consistent with this, none of the top five mean-MR-ranked candidates -- two inositol-3-phosphate synthase paralogs, a katanin p60 ATPase subunit, a DEAD-box RNA helicase, and achilleol B synthase -- carry a functional annotation connected to the TCA cycle or organic-acid metabolism, the pathway aconitate isomerase belongs to. This does not rule any of them out: enzyme-activity annotation in a non-model crop genome is itself incomplete, so a true aconitate isomerase could plausibly be mis-annotated under a different enzyme family. But it also means these candidates do not, on their own, tell an obvious mechanistic story the way a HIGH-impact coding variant overlapping a known dPCD executor gene would. They are best read as a prioritised hypothesis space for targeted follow-up (e.g.\ heterologous expression or knockout validation of the top-ranked genes) rather than as a confident nomination.

% \hspace{1em}Taken together, the two case studies show that the platform can connect multi-omics evidence to candidate gene lists, and that each layer can produce a real, checkable signal when its underlying assumption matches the phenotype's true molecular mechanism. The D-gene case study also shows the reverse: a layer can fail cleanly and informatively when that assumption does not hold, as when a genetic loss-of-function mechanism is searched for with an epigenetic-silencing hypothesis, or when a regulatory gene is searched for within a module curated around its downstream executors. Applied to TAA, where no positive control exists to calibrate these assumptions against, the resulting candidate lists should be treated as the platform is designed to produce them: prioritised, testable hypotheses rather than final identifications.
